Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
updates
More updates

\begin{abstract}
Effect handlers have been gathering momentum as a mechanism for modular
	programming with user-defined effects. Effect handlers allow for non-local
	control flow mechanisms such as generators, async/await, lightweight threads
	and coroutines to be composably expressed.  We present a design and evaluate a
	full-fledged efficient implementation of effect handlers for OCaml, an
	industrial-strength multi-paradigm programming language. Our implementation
	strives to maintain the backwards compatibility and performance profile of
	existing OCaml code. Retrofitting effect handlers onto OCaml is challenging
	since OCaml does not currently have any non-local control flow mechanisms
	other than exceptions. Our implementation of effect handlers for OCaml: {\em
	(i)} imposes negligible overhead on code that does not use effect handlers;
	{\em (ii)} remains compatible with program analysis tools that inspect the
	stack; and {\em (iii)} is efficient for new code that makes use of effect
	handlers.
\end{abstract}

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
		better than legacy?} \tk{Existing might be better?} do not break under
    OCaml extended with effect handlers. OCaml code that does not use effect
    handlers will pay negligible performance and memory cost.

@@

\begin{itemize}
	\item the design and implementation of effect handlers for OCaml. Our design
		retains OCaml's compatibility with program analysis tools that inspect the
		stack using DWARF unwind tables. We have validated our DWARF unwind tables
		with the assistance of an automated validator tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

Figure~\ref{res:macro_time} presents the running time of the different
multicore variants normalized against the sequential baseline |stock|. On
average (geometric mean of the normalized values against |stock| as the
baseline), the multicore variants were \textbf{less than 1\% slower} than
|stock|. The outliers (on either ends) were due to the difference in the
allocator and the GC between stock and Multicore OCaml.

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
\vspace{-5mm}
}
\label{tab:micro_noeffect}
\end{table}

@@

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and negligible overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of effect handlers into OCaml
implemented using lightweight fibers, along with a parallel
runtime~\cite{Sivaramakrishnan20}, as has been demonstrated in our work, will
open OCaml to highly scalable concurrent and task-parallel applications with
minimal hit to sequential performance.

%\clearpage

>>>>

\begin{abstract}
Effect handlers have been gathering momentum as a mechanism for modular
	programming with user-defined effects. Effect handlers allow for non-local
	control flow mechanisms such as generators, async/await, lightweight threads
	and coroutines to be composably expressed.  We present a design and evaluate
	a full-fledged efficient implementation of effect handlers for OCaml, an
	industrial-strength multi-paradigm programming language. Our implementation
	strives to maintain the backwards compatibility and performance profile of
	existing OCaml code. Retrofitting effect handlers onto OCaml is challenging
	since OCaml does not currently have any non-local control flow mechanisms
	other than exceptions. Our implementation of effect handlers for OCaml: {\em
	(i)} imposes a mean 1\% overhead on a comprehensive macro benchmark suite
	that does not use effect handlers; {\em (ii)} remains compatible with program
	analysis tools that inspect the stack; and {\em (iii)} is efficient for new
	code that makes use of effect handlers.
\end{abstract}

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} Existing OCaml programs do not break
		under OCaml extended with effect handlers. OCaml code that does not use
		effect handlers will pay minimal performance and memory cost.

@@

\begin{itemize}
	\item the design and implementation of effect handlers for OCaml. Our design
		retains OCaml's compatibility with program analysis tools that inspect the
		stack using DWARF unwind tables. We have validated our DWARF unwind tables
		with the assistance of an automated validator tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has minimal
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

Figure~\ref{res:macro_time} presents the running time of the different
multicore variants normalized against the sequential baseline |stock|. On
average (geometric mean of the normalized values against |stock| as the
baseline), the multicore variants were less than 1\% slower than |stock|. The
outliers (on either ends) were due to the difference in the allocator and the
GC between stock and Multicore OCaml. Of the 54 programs in the benchmark
suite, 32 programs had an overhead of 5\% or lower, and 8 programs had more
than 10\% overhead.

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
}
\label{tab:micro_noeffect}
\vspace{-3mm}
\end{table}

@@

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and minimal overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of effect handlers into OCaml
implemented using lightweight fibers, along with a parallel
runtime~\cite{Sivaramakrishnan20}, as has been demonstrated in our work, will
open OCaml to highly scalable concurrent and task-parallel applications with
minimal hit to sequential performance.

\section*{Acknowledgements}

This research was funded via a Royal Commission for the Exhibition
of 1851 and Darwin College Research Fellowships, and by grants from Jane Street
and Tezos Foundation.
